Supplementary Information

### 2020-11-27 doc/manuscript/MEE_manuscript_SI.tex
worked on manuscript

@@

For each core, \(DBH\) can be reconstructed outside-in (based on recent
\(DBH\), subtracting growth recorded in tree rings) or inside-out
(summing \(RW\) from the inside out). We generally gave precedence to
the outside-in approach. Specifically, when \(DBH\) was taken at the
time of coring,\\
At some of our sites where DBH was not taken at the time of coring
(\emph{SCBI},), DBH measurements taken before or slightly after the time
of coring could be used. (see
\href{https://github.com/EcoClimLab/ForestGEO_dendro/issues/19}{issue
\#19 in ForestGEO\_dendro}) If before, \ldots{} If after\ldots{} For all
outside-in reconstructions, if a negative \(DBH\) was predicted\ldots{}

@@

\hypertarget{appendix-s2.-methods-for-comparing-climwin-results-with-traditional-methods}{%
\subsection{Appendix S2. Methods for comparing climwin results with
traditional
methods}\label{appendix-s2.-methods-for-comparing-climwin-results-with-traditional-methods}}

\emph{(\href{https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35}{ISSUE
\#35 in ForestGEO-climate-sensitivity})}

To verify that our methods gave similar results to traditional methods,
we conducted qualitative comparisons of our results to previous studies
based on the same cores (Table \textbf{S5}). We also conducted a formal
comparison using identical tree-ring and climate data for four
well-studied species: PSME (Cedar Breaks, Utah), ABAL (Zofin), PIMA
(Scotty Creek), and LITU (SCBI) (Fig. \textbf{S1}). We compared results
from an analysis using conventional methods, as detailed below, to an
analysis using our method as described in the Methods section, but with
the \emph{climwin} climate variable selection process limited to just
the species of interest (as opposed to all species at the site), climate
variables considered individually rather than additively, and with start
date adjusted to match the conventional method.

The ring-width series from each core was standardized via ARSTAN using a
2/3rds \(n\) spline, where \(n\) is the number of years in the series
(\emph{Cook, 1985; Cook \& Kairiukstis, 1990- citations in Helcoski}).
\emph{(The following italic text is plagarized from Helcoski and needs
to be reworded:)} \emph{The influence of outliers in all series was
reduced using the adaptive power transformation, which also stabilises
the variance over time (Cook \& Peters, 1997). Next, each series was
stabilised using either the average correlation between raw ring-width
series (rbar) method or a 1/ 3rds spline method to adjust changes in
variance as series replication decreased towards the earlier portion of
each chronology (Jones et al., 1997). The 1/3rds spline method was
chosen when replication in the inner portion of each chronology (c.~the
inner 30--50 yr of each record depending on full chronology length)
dropped below three trees. Once that step was complete, a robust
biweight mean chronology for each species was calculated from the
ring-width indices (Cook, 1985). We chose to use residual chronologies
because the autoregressive standardisation process in creating them
removes much of the tree-level autocorrelation in growth and these
chronologies would most likely contain the most conservative information
on drivers of interannual growth (Cook, 1985).}

Following Helcoski et al. (2019), we defined chronology start dates
according to the subsample signal strength (SSS), using a cutoff of SSS
= 0.80 (or 80\% of the population signal). Thus, for this analysis only,
we defined chronology start dates as the year the SSS exceeded 0.80 or
two years after the start of the climate record, whichever came later.
SSS exceeded 0.80 well before the start of the 1901 start of climate
records for PSME (1800s), ABAL (1700), and PIMA (1850s). For LITU , SSS
reached 0.8 with 11 trees in 1919, which we used as the start date for
this series. We note that these start date criteria differ from those
used in the main analysis (Table S3), which had earlier start dates
because the analysis was not constrained by a need to represent the full
population signal. End dates were defined as the last full year prior to
sampling (Table S3).

@@

\hypertarget{appendix-s3.-dealing-with-rapidly-changing-climate-and-tree-growth}{%
\subsection{Appendix S3. Dealing with rapidly changing climate and tree
growth}\label{appendix-s3.-dealing-with-rapidly-changing-climate-and-tree-growth}}

@@

Our analysis included two sites where climate change has had pronounced
effects on tree growth: Scotty Creek, NW Territories, Canada (SC) and
Little Tesuque, New Mexico, USA (LT). At SC, {[}temperatures have
increased by X \(^\circ\) over X years{]}\ldots, resulting in negative
growth trends in basal area index (\(BAI\)) starting around 1950 and
significant growth declines since 1970 in 56\% of trees (Sniderhan \&
Baltzer, 2016). At LT, \emph{(drought has increased dramatically)},
resulting in many missing rings in recent years.

@@

\hypertarget{table-s1.-site-details}{%
\subsection{Table S1. Site Details}\label{table-s1.-site-details}}

\begin{table}[!h]
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}{l>{\raggedright\arraybackslash}p{3cm}rr>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{1.5cm}llll}
\toprule
site code & site name & latitude & longitude & elevation (m.a.s.l.) & cores within ForestGEO plot? & canopy positions & tree statuses & date range & dormant season & months in climwin\\
\midrule
BCI & Barro Colorado Island & 9.15430 & -79.8461 & 120-160 & no & canopy & live, dead & 1931-2014 & Nov-Apr & pOct-cDec\\
HKK & Huai Kha Khaeng & 15.63240 & 99.2170 & 549-638 & no & all & live & 1903-2011 & Nov-Apr & pOct-cDec\\
LT & Little Tesuque & 35.73838 & -105.8382 & 2682 & n.a. & all & live & 1903-2018 &  & pMay-cAug\\
CB & Utah Forest Dynamics Plot & 37.66150 & -112.8525 & 3020-3169 & yes &  & live & 1903-2007 &  & pMay-cAug\\
SCBI & Smithsonian Conservation Biology Institute & 38.89350 & -78.1454 & 273-338 & yes & all & live, dead & 1903-2017 & Oct-Apr & pMay-cAug\\
\addlinespace
LDW & Lilly Dickey Woods & 39.23590 & -86.2181 & 230-303 &  & canopy & live, dead & 1903-2019 &  & pMay-cAug\\
HF & Harvard Forest & 42.53880 & -72.1755 & 340-368 & yes & all & live, dead & 1903-2014 &  & pMay-cAug\\
NE & Niobrara/Halsey & 42.78000 & -100.0210 & 644-702 & some & canopy & live &  & Oct-Apr & pMay-cAug\\
ZOF & $\v{Z}$of$\'{i}$n Forest Dynamics Plot & 48.66380 & 14.7073 & 736-829 & some & all & live, dead & 1903-2013 & Oct-Mar & pMay-cAug\\
SC & Scotty Creek & 61.30000 & -121.3000 & 258-274 & no & all & live, dead & 1903-2013 &  & pMay-cAug\\
\bottomrule
\end{tabular}}
\end{table}

@@

\hypertarget{table-s2.-species-analyzed-their-characteristics-and-bark-allometries-applied}{%
\subsection{Table S2. Species analyzed, their characteristics, and bark
allometries
applied}\label{table-s2.-species-analyzed-their-characteristics-and-bark-allometries-applied}}

@@

ABAL & Pinaceae & Abies alba & ZOF & needleleaf & evergreen & shade-tolerant & neglected in Zofin\\

@@

ABBI & Pinaceae & Abies bifolia & CB & needleleaf & evergreen &  & neglected in CedarBreaks\\

@@

FRAM & Oleaceae & Fraxinus americana & LDW, SCBI & broadleaf & deciduous (cold) &  & Fraxinus ssp. in LillyDickey, fram in SCBI\\

@@

JUVI & Cupressaceae & Juniperus virginiana & NE &  &  &  & neglected in Nebraska\\

@@

PIAB & Pinaceae & Picea abies & HF & needleleaf & evergreen & intermediate & neglected in HarvardForest, neglected in Zofin\\

@@

PIPU & Pinaceae & Picea pungens & CB & needleleaf & evergreen &  & neglected in CedarBreaks\\

@@

PIST & Pinaceae & Pinus strobus & HF, SCBI & needleleaf & evergreen &  & Pinus strobus in HarvardForest, pist in SCBI\\
PIST2 & Pinaceae & Pinus strobiformis & LT & needleleaf & evergreen &  & Pinus monticola in Little Tesuque\\

@@

QURU & Fagaceae & Quercus rubra & HF, LDW, SCBI & broadleaf & deciduous (cold) &  & quru in HarvardForest, Quercus rubra in LillyDickey, quru in SCBI\\

@@

\hypertarget{table-s3.-sampling-details-for-species-by-site}{%
\subsection{Table S3. Sampling details for species by
site}\label{table-s3.-sampling-details-for-species-by-site}}

@@

BCI & JACO & 12 & 18 & 11 & 17 & 30.2-63.5 & 2.6-56.4 & 1931-2014\\

@@

BCI & TEPA & 18 & 29 & 17 & 26 & 22.1-59.5 & 2.7-49.4 & 1931-2014\\

@@

BCI & TRTU & 23 & 37 & 20 & 31 & 20.7-43.6 & 4.8-41.5 & 1931-2014\\

@@

CB & ABBI & 22 & 41 & 20 & 37 & 13.9-54.2 & 0.9-46.4 & 1903-2000\\

@@

CB & PIEN & 12 & 21 & 10 & 15 & 14-54.9 & 0.9-33.1 & 1903-2000\\

@@

CB & PIFL & 13 & 22 & 12 & 21 & 17.6-64.1 & 1.5-47.5 & 1903-1998\\

@@

CB & PILO & 11 & 16 & 8 & 12 & 45.8-121.6 & 35.4-118.5 & 1903-1999\\

@@

CB & PIPU & 15 & 28 & 15 & 28 & 22.4-50.8 & 9.5-48.4 & 1903-2000\\

@@

CB & POTR & 17 & 27 & 17 & 26 & 23.6-47.6 & 4.3-35.4 & 1903-2000\\

@@

CB & PSME & 11 & 20 & 10 & 18 & 20.7-64.2 & 0.5-41.5 & 1903-1999\\

@@

HF & ACRU & 18 & 59 & 18 & 59 & 10.1-22.1 & 0.9-20.4 & 1903-2013\\

@@

HF & BEAL & 13 & 44 & 13 & 44 & 10.2-37.9 & 0-17.2 & 1904-2013\\

@@

HF & QURU & 74 & 180 & 73 & 177 & 19.5-53 & 1.1-48.3 & 1903-2014\\

@@

HF & TSCA & 32 & 83 & 32 & 71 & 10.6-37 & 0-28.4 & 1923-2014\\

@@

HKK & AFXY & 39 & 127 & 39 & 127 & 20.1-98.7 & 0.1-81.4 & 1903-2011\\

@@

HKK & CHTA & 28 & 70 & 28 & 70 & 16-64.6 & 0.2-59.5 & 1904-2010\\

@@

HKK & MEAZ & 46 & 130 & 46 & 130 & 25.6-98.1 & 3.8-80.3 & 1914-2011\\

@@

HKK & TOCI & 45 & 143 & 45 & 143 & 16.6-116.4 & 1.7-80.5 & 1903-2011\\

@@

LDW & ACSA & 35 & 66 & 34 & 64 & 9-64.6 & 0-52.4 & 1903-2019\\

@@

LDW & LITU & 15 & 28 & 14 & 25 & NA-NA & 1.2-69.4 & 1903-2019\\

@@

NE & BEPA & 28 & 84 & 28 & 84 & NA-NA & 0-25.5 & 1903-1995\\

@@

NE & JUVI & 29 & 60 & 29 & 60 & 16.6-21.5 & 0.3-18.4 & 1938-1994\\

@@

NE & PIPO & 71 & 134 & 70 & 129 & 19.3-64 & 0-41.1 & 1934-1994\\

@@

LT & PIPO & 10 & 20 & 10 & 20 & 23.2-52.8 & 14.4-48.4 & 1903-2018\\

@@

LT & PIST2 & 7 & 14 & 7 & 14 & 25.7-39.8 & 3.9-34.5 & 1903-2018\\

@@

SCBI & CACO & 15 & 15 & 14 & 14 & 10.62-38.52 & 1.7-32.2 & 1903-2015\\

@@

SCBI & CAGL & 39 & 39 & 36 & 36 & 10.28-52.31 & 1.6-49.3 & 1903-2015\\

@@

SCBI & CAOVL & 25 & 25 & 24 & 24 & 15.11-60.32 & 2.6-47.2 & 1903-2015\\

@@

SCBI & CATO & 15 & 15 & 14 & 14 & 12.86-35.95 & 3.7-28.4 & 1903-2015\\

@@

SCBI & FAGR & 76 & 76 & 74 & 74 & 10.05-41.02 & 0.1-41.2 & 1920-2009\\

@@

SCBI & FRAM & 66 & 66 & 61 & 61 & 8.11-94.73 & 0.1-84.4 & 1903-2016\\

@@

SCBI & FRNI & 12 & 12 & 12 & 12 & 11.04-39.2 & 0.5-27.3 & 1903-1996\\

@@

SCBI & JUNI & 30 & 30 & 28 & 28 & 20.4-76.19 & 5.6-59.5 & 1903-2010\\

@@

SCBI & LITU & 106 & 106 & 104 & 104 & 10-91.42 & 0.1-81.1 & 1903-2010\\

@@

SCBI & PIST & 36 & 36 & 36 & 36 & 13.92-50.96 & 1.6-45.2 & 1931-2010\\

@@

SCBI & QUAL & 66 & 66 & 66 & 66 & 11.4-76.73 & 0.3-70.4 & 1903-2009\\

@@

SCBI & QUMO & 67 & 67 & 67 & 67 & 10.22-84.59 & 0.3-69.5 & 1903-2017\\

@@

SCBI & QURU & 70 & 70 & 70 & 70 & 11.07-87.65 & 2.5-79.2 & 1903-2016\\

@@

SCBI & QUVE & 81 & 81 & 81 & 81 & 16.02-82.33 & 0.5-78.4 & 1903-2009\\

@@

SC & PIMA & 442 & 442 & 101 & 101 & 7-14.9 & 0.5-12.5 & 1903-2013\\

@@

ZOF & ABAL & 55 & 55 & 50 & 50 & 41-121 & 21.3-108.5 & 1903-2011\\

@@

ZOF & FASY & 1369 & 1369 & 1358 & 1358 & NA-NA & 0.1-115.3 & 1903-2013\\

@@

ZOF & PCAB & 644 & 644 & 599 & 599 & NA-NA & 0.5-126.4 & 1903-2011\\

@@

\hypertarget{table-s4.-allometric-equations-for-bark-thickness}{%
\subsection{Table S4. Allometric equations for bark
thickness}\label{table-s4.-allometric-equations-for-bark-thickness}}

@@

\em{Acer rubrum} & $bark.mm = 0.619 * log(dbh.cm + 1)$ & 10 & 8.2-39.6 & SCBI & Anderson-Teixeira et al. (2015)\\

@@

\em{Betula alleghaniensis} & $bark.mm = (0.15 + 0.03 * dbh.cm)/2$ & NA &  & North America average of its range & Miles, Patrick D.; Smith, W. Brad. 2009.\\

@@

@@

\em{Carya ovata} & $bark.mm = 1.035 * log(dbh.cm + 1)$ & 8 & 19.1-78 & SCBI & Anderson-Teixeira et al. (2015)\\

>>>>

\hypertarget{barro-colorado-island-panama}{%
\subsubsection{Barro Colorado Island,
Panama}\label{barro-colorado-island-panama}}

\hypertarget{appendix-s2.-methods-for-reconstruction-of-dbh}{%
\subsection{\texorpdfstring{Appendix S2. Methods for reconstruction of
\(DBH\)}{Appendix S2. Methods for reconstruction of DBH}}\label{appendix-s2.-methods-for-reconstruction-of-dbh}}

@@

In most cases, when a recent \(DBH\) measurement was available, \(DBH\)
was reconstructed from the outside in. In cases where \(DBH\) was not
available, but when we knew that the core hit pith or could reasonably
estimate how far off it was based on the curvature of the rings
(Applequist, 1958; Duncan, 1989), \(DBH\) was reconstructed from the
inside out.

For each core, \(DBH\) can be reconstructed outside-in (based on recent
\(DBH\), subtracting growth recorded in tree rings) or inside-out
(summing \(RW\) from the inside out,--only when core hit pith or
distance to pith can be reliably estimated). We generally gave
precedence to the outside-in approach. Specifically, when \(DBH\) was
taken at the time of coring,\\
At some of our sites where DBH was not taken at the time of coring
(\emph{SCBI},), DBH measurements taken before or slightly after the time
of coring could be used. (see
\href{https://github.com/EcoClimLab/ForestGEO_dendro/issues/19}{issue
\#19 in ForestGEO\_dendro}) If before, \ldots{} If after\ldots{} For all
outside-in reconstructions, if a negative \(DBH\) was predicted\ldots{}

@@

\hypertarget{appendix-s3.-methods-for-climate-data-evaluation-and-correction}{%
\subsection{Appendix S3. Methods for climate data evaluation and
correction}\label{appendix-s3.-methods-for-climate-data-evaluation-and-correction}}

\emph{For BCI, we calculated monthly \(PPT\) and \(PDF\) from daily
precipitation readings made on BCI starting in 1929 (Paton, 2019).}

@@

\hypertarget{appendix-s4.-methods-for-comparing-climwin-results-with-traditional-methods}{%
\subsection{Appendix S4. Methods for comparing climwin results with
traditional
methods}\label{appendix-s4.-methods-for-comparing-climwin-results-with-traditional-methods}}

To test whether our methods gave similar results to traditional methods,
we conducted qualitative comparisons of our results to previous studies
based on the same cores (Table S5) and conducted a formal quantitative
comparison for four species (Figs. S1-S4), as detailed below.

\emph{Qualitative comparison}

For all species-site combinations, we reviewed previous studies
characterizing the climate sensitivity of growth using conventional
methods. In most cases, we were able to compare with previous studies
from the same sites and sets of cores. When these were not available, we
reviewed regional-level analyses believed to be representative of the
site.

Results from previous studies were compiled alongside results from the
climate-only model in this study (Table S5). Where previous studies
examined numerous climate variables or time windows (e.g., Helcoski et
al., 2019), we focus on those most relevant to our findings.

Beyond the methodological differences, original studies based on the
same sets of cores varied from this one and from one another in factors
including the exact set of cores analyzed, climate data sources, time
frame of analysis, approaches to identifying candidate climate variables
and windows (including whether this is done on a site or species level),
methods for detrending and standardizing to build chronologies, and
whether the effects of temperature and precipitation are considered
separately (original studies) or additively (this study). To standardize
for such differences, we selected a subset of species for a standardized
quantitative comparison, as detailed below.

\emph{Quantitative comparison}

We also conducted a formal comparison of our approach to conventional
methods using identical tree-ring and climate data for four species:
PSME (Cedar Breaks, Utah), ABAL (\(\v{Z}\)of\(\'{i}\)n), PIMA (Scotty
Creek), and LITU (SCBI; Figs. S1-S4). These species were selected for
analysis because they have been well-studied in the past. For each
species, we compared climate sensitivities for the top precipitation-
and temperature- group variables, as identified in the main analysis.

Prior to analysis, data were prepared and cleaned as described in the
Methods section, resulting in an identical set of records for input into
each analysis. For the approach developed here, analysis was conducted
as described in the Methods section, but with the \emph{climwin} climate
variable selection process limited to just the species of interest (as
opposed to all species at the site), climate variables considered
individually rather than additively, analysis of only first-order linear
relationships, and with start date adjusted to match the conventional
method (see below). Following the \emph{climwin} analysis step, we
extracted \emph{beta} coefficients describing the slope of the
relationship between climate and \(RW\). \emph{Beta} coefficients, along
with their standard error, were obtained for each month within the
analysis time frame (Table S1) and for the time window identified as
optimal by \emph{climwin}.

For the analysis using conventional methods, the ring-width series from
each core was standardized via ARSTAN using a 2/3rds \(n\) spline, where
\(n\) is the number of years in the series (\emph{Cook, 1985; Cook \&
Kairiukstis, 1990- citations in Helcoski}). \emph{(The following italic
text is self-plagarized from Helcoski and needs to be reworded:)}
\emph{The influence of outliers in all series was reduced using the
adaptive power transformation, which also stabilises the variance over
time (Cook \& Peters, 1997). Next, each series was stabilised using
either the average correlation between raw ring-width series (rbar)
method or a 1/ 3rds spline method to adjust changes in variance as
series replication decreased towards the earlier portion of each
chronology (Jones et al., 1997). The 1/3rds spline method was chosen
when replication in the inner portion of each chronology (c.~the inner
30--50 yr of each record depending on full chronology length) dropped
below three trees. Once that step was complete, a robust biweight mean
chronology for each species was calculated from the ring-width indices
(Cook, 1985). We chose to use residual chronologies because the
autoregressive standardisation process in creating them removes much of
the tree-level autocorrelation in growth and these chronologies would
most likely contain the most conservative information on drivers of
interannual growth (Cook, 1985).}

Following Helcoski et al. (2019), we defined chronology start dates
according to the subsample signal strength (SSS), using a cutoff of SSS
= 0.80 (or 80\% of the population signal). Thus, for this analysis only,
we defined chronology start dates as the year the SSS exceeded 0.80 or
two years after the start of the climate record, whichever came later.
SSS exceeded 0.80 well before the start of the 1901 start of climate
records for PSME (1800s), ABAL (1700), and PIMA (1850s). For LITU, SSS
reached 0.8 with 11 trees in 1919, which we used as the start date for
this series. We note that these start date criteria differ from those
used in the main analysis (Table S3), which had earlier start dates
because the analysis was not constrained by a need to represent the full
population signal. End dates were defined as the last full year prior to
sampling (Table S3).

\emph{Beta} (slope) coefficients for the relationship between tree
growth and the monthly climate variable were derived as in Helcoski et
al. (2019): \emph{(SELF-PLAGIARIZED CONTENT:) Analyses of
climate--growth relationships were conducted using `dplR' (Bunn, 2008)
and `bootRes' (Zang \& Biondi, 2013), which correlated functions and
bootstrapped confidence intervals for the relationships between annual
growth and monthly climate variables following Biondi \& Waikul (2004).}
Pearson correlations between climate variables and tree-ring
chronologies were converted to linear slopes using the method of Charney
et al. (2016).

Finally, we generated plots comparing month-by-month \emph{beta}
coefficients describing climate sensitivity, and also comparing
\emph{beta} coefficients for the window identified as optimal by
\emph{climwin} Figs. S1-S4).

We note that despite designing the analyses to be as comparable as
possible, one-to-one correspondence of \emph{beta} coefficients is not
necessarily expected for several reasons. First, although the analysis
time frame is standardized between the two approaches, the relative
influence of each year will generally vary between the two approaches.
The traditional approach, which all cores into a single residual
chronology with one value per year, gives equal weighting to each year.
In contrast, under the approach developed here, the number records per
year will vary across the analysis time frame, generally increasing over
time as the younger trees enter the analysis. Thus, where many younger
trees are included in the analysis, the two approaches will effectively
give different weights to the years included in the analysis period. In
cases where climate-sensitivity differs between old and young trees, or
where the climate and/or climate response changed substantially over the
analysis time frame (e.g., at Scotty Creek; Fig. S4; Sniderhan \&
Baltzer, 2016), this may lead to divergence of the climate sensitivities
estimated by the two methods.

Second, traditional analysis methods (using ARSTAN) were primarily
designed to distill population-level variation to obtain the strongest
possible climate signal for the reconstruction of past climate
(\textbf{Cook \& Kariustis }), not to characterize climate responses on
the individual level, where variation is inherently higher. While
conversion of Pearson correlations to linear slopes \emph{sensu} Charney
et al. (2016) approximates climate responses, it does not provide an
exact slope describing the relationship between individual-level or
population mean growth and climate. This is because standardization of
variance and averaging of individual-level residuals prior to the
climate analysis fundamentally alters and obfuscates individual-level
responses.

We suspect that both of these factors may underlie the tendency for the
traditional method to estimate stronger climate sensitivity than the
approach developed here for Scotty Creek (Fig. S4), a comprehensively
sampled black spruce forest (i.e., including young trees) on melting
permafrost. We note, however, that there are no statistically
significant differences in the \emph{beta} coefficients of the two
approaches at this site.

\hypertarget{appendix-s5.-dealing-with-rapidly-changing-climate-and-tree-growth}{%
\subsection{Appendix S5. Dealing with rapidly changing climate and tree
growth}\label{appendix-s5.-dealing-with-rapidly-changing-climate-and-tree-growth}}

@@

Our analysis included two sites where climate change has had pronounced
effects on tree growth: Scotty Creek, NW Territories, Canada (SC) and
Little Tesuque, New Mexico, USA (LT). At SC, rapidly rising temperatures
are causing melting permafrost, summer moisture stress, resulting in
negative growth trends in basal area index (\(BAI\)) starting around
1950 and significant growth declines since 1970 in 56\% of trees
(Sniderhan \& Baltzer, 2016). At LT, increasingly warm drought has
dramatically reduced growth (Williams et al., 2012), resulting in many
missing rings in recent years.

Problematically, correlating tree growth residuals from which
climate-driven trends had been removed against the climate signal with a
strong directional trend would not necessarily identify the most
relevant climate drivers.

For these sites, we experimented with three approaches to identifying
the most important climate drivers (1) the method described above, (2)
detrending the climate variables \textbf{(AT:prewhitening?)} prior to
the climwin step, and (3) splitting analyses into decades before and
after 1970 (\emph{sensu} Sniderhan \& Baltzer, 2016) (Appendix S5).

@@

\hypertarget{table-s1.-site-details.}{%
\subsection{Table S1. Site Details.}\label{table-s1.-site-details.}}

\begin{table}[!h]
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}{l>{\raggedright\arraybackslash}p{3cm}rr>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{1.5cm}llll}
\toprule
site code & site name & latitude & longitude & elevation (m.a.s.l.) & cores within ForestGEO plot? & canopy positions & tree statuses & date range & dormant season* & months in climwin\\
\midrule
BCI & Barro Colorado Island & 9.15430 & -79.8461 & 120-160 & no & canopy & live, dead & 1931-2014 & Nov-Apr & pOct-cDec\\
HKK & Huai Kha Khaeng & 15.63240 & 99.2170 & 549-638 & no & all & live & 1903-2011 & Nov-Apr & pOct-cDec\\
SCBI & Smithsonian Conservation Biology Institute & 38.89350 & -78.1454 & 273-338 & yes & all & live, dead & 1903-2017 & Oct-Apr & pMay-cAug\\
LDW & Lilly Dickey Woods & 39.23590 & -86.2181 & 230-303 &  & canopy & live, dead & 1903-2019 &  & pMay-cAug\\
HF & Harvard Forest & 42.53880 & -72.1755 & 340-368 & yes & all & live, dead & 1903-2014 &  & pMay-cAug\\
\addlinespace
ZOF & $\v{Z}$of$\'{i}$n Forest Dynamics Plot & 48.66380 & 14.7073 & 736-829 & some & all & live, dead & 1903-2013 & Oct-Mar & pMay-cAug\\
NE & Niobrara/Halsey & 42.78000 & -100.0210 & 644-702 & some & canopy & live &  & Oct-Apr & pMay-cAug\\
LT & Little Tesuque & 35.73838 & -105.8382 & 2684 - 2702 & n.a. & canopy/ subcanopy & live & 1903-2018 & Oct-Apr & pMay-cAug\\
CB & Utah Forest Dynamics Plot & 37.66150 & -112.8525 & 3020-3169 & yes &  & live & 1903-2007 &  & pMay-cAug\\
SC & Scotty Creek & 61.30000 & -121.3000 & 280 & no & all & live, dead & 1903-2013 & Sept-Apr & pMay-cAug\\
\bottomrule
\end{tabular}}
\end{table}

\begin{itemize}
\tightlist
\item
  Refers to approximate period during which woody growth ceases (dry
  season in the tropics, winter for temperate and boreal sites).
\end{itemize}

@@

\hypertarget{table-s2.-species-analyzed-their-characteristics-and-bark-allometries-applied.}{%
\subsection{Table S2. Species analyzed, their characteristics, and bark
allometries
applied.}\label{table-s2.-species-analyzed-their-characteristics-and-bark-allometries-applied.}}

@@

ABAL & Pinaceae & Abies alba & ZOF & needleleaf & evergreen & shade-tolerant & Abies lasiocarpa in Zofin\\

@@

ABBI & Pinaceae & Abies bifolia & CB & needleleaf & evergreen &  & Abies lasiocarpa in CedarBreaks\\

@@

FRAM & Oleaceae & Fraxinus americana & LDW, SCBI & broadleaf & deciduous (cold) &  & fram in LillyDickey, fram in SCBI\\

@@

JUVI & Cupressaceae & Juniperus virginiana & NE & needleleaf & evergreen &  & neglected in Nebraska\\

@@

PIAB & Pinaceae & Picea abies & HF & needleleaf & evergreen & intermediate & Picea engelmannii in HarvardForest, Picea engelmannii in Zofin\\

@@

PIPU & Pinaceae & Picea pungens & CB & needleleaf & evergreen &  & Picea engelmannii in CedarBreaks\\

@@

PIST & Pinaceae & Pinus strobus & HF, SCBI & needleleaf & evergreen &  & Pinus strobus in HarvardForest, Pinus strobus in SCBI\\

@@

QURU & Fagaceae & Quercus rubra & HF, LDW, SCBI & broadleaf & deciduous (cold) &  & quru in HarvardForest, quru in LillyDickey, quru in SCBI\\

@@

\hypertarget{table-s3.-sampling-details-for-species-by-site.}{%
\subsection{Table S3. Sampling details for species by
site.}\label{table-s3.-sampling-details-for-species-by-site.}}

@@

BCI & JACO & 12 & 18 & 11 & 17 & 30.2-63.5 & 2.6-56.4 & 1937-2014\\

@@

BCI & TEPA & 18 & 29 & 17 & 26 & 22.1-59.5 & 3-49.4 & 1937-2014\\

@@

BCI & TRTU & 23 & 36 & 20 & 31 & 20.7-43.6 & 4.8-41.5 & 1937-2014\\

@@

CB & ABBI & 22 & 41 & 20 & 37 & 13.9-54.2 & 0.6-45.3 & 1903-1999\\

@@

CB & PIEN & 12 & 21 & 10 & 16 & 14-54.9 & 3.6-39.3 & 1903-1999\\

@@

CB & PIFL & 12 & 19 & 11 & 18 & 17.6-63.5 & 4.5-55.4 & 1903-1998\\

@@

CB & PILO & 17 & 24 & NA & NA & NA & NA & 1903-1999\\

@@

CB & PIPU & 15 & 28 & 15 & 28 & 22.4-50.8 & 8.6-48.5 & 1903-1999\\

@@

CB & POTR & 17 & 27 & 17 & 27 & 23.6-47.6 & 7.7-42.4 & 1903-1999\\

@@

CB & PSME & 10 & 19 & 10 & 19 & 20.7-64.2 & 2.6-49.4 & 1903-1999\\

@@

HF & ACRU & 18 & 59 & 18 & 59 & 10.1-22.1 & 1.7-20.4 & 1932-2013\\

@@

HF & BEAL & 12 & 41 & 12 & 41 & 10.2-21.1 & 1.6-20.5 & 1932-2013\\

@@

HF & QURU & 74 & 179 & 73 & 176 & 19.5-53 & 1.4-48.3 & 1932-2013\\

@@

HF & TSCA & 32 & 83 & 32 & 82 & 10.6-37 & 0.6-33.4 & 1932-2013\\

@@

HKK & AFXY & 37 & 115 & 37 & 115 & 20.1-80.8 & 1.2-80.8 & 1964-2010\\

@@

HKK & CHTA & 27 & 68 & 27 & 68 & 16-59.1 & 1.1-59.1 & 1964-2010\\

@@

HKK & MEAZ & 45 & 121 & 45 & 121 & 25.6-83.2 & 3.8-79.6 & 1964-2010\\

@@

HKK & TOCI & 44 & 138 & 44 & 138 & 16.6-80.3 & 2-78.4 & 1964-2010\\

@@

LDW & ACSA & 35 & 66 & 34 & 64 & 9-64.6 & 0-52.4 & 1903-2013\\

@@

LDW & LITU & 15 & 28 & 14 & 25 & NA-NA & 1.2-69.4 & 1903-2013\\

@@

NE & BEPA & 28 & 82 & 28 & 82 & NA-NA & 0.4-30.5 & 1946-1994\\

@@

NE & JUVI & 29 & 60 & 29 & 60 & 16.6-21.5 & 1.8-18.4 & 1946-1994\\

@@

NE & PIPO & 59 & 113 & 57 & 109 & 19.3-63 & 2.2-46.9 & 1946-1994\\

@@

LT & PIPO & 10 & 20 & 10 & 20 & 23.2-52.8 & 14.6-48.4 & 1903-2018\\

@@

LT & PIST2 & 7 & 14 & 7 & 13 & 25.7-39.8 & 4.2-34.4 & 1903-2018\\

@@

SCBI & CACO & 15 & 15 & 14 & 14 & 10.62-38.52 & 1.7-32.2 & 1932-2009\\

@@

SCBI & CAGL & 39 & 39 & 36 & 36 & 10.28-52.31 & 1.9-49.3 & 1932-2009\\

@@

SCBI & CAOVL & 25 & 25 & 24 & 24 & 15.11-60.32 & 4.7-47.2 & 1932-2009\\

@@

SCBI & CATO & 15 & 15 & 14 & 14 & 12.86-35.95 & 3.7-28.4 & 1932-2009\\

@@

SCBI & FAGR & 76 & 76 & 74 & 74 & 10.05-41.02 & 0.1-41.2 & 1932-2009\\

@@

SCBI & FRAM & 66 & 66 & 61 & 61 & 8.11-94.73 & 0.1-84.4 & 1932-2009\\

@@

SCBI & FRNI & 12 & 12 & 12 & 12 & 11.04-39.2 & 0.7-27.3 & 1932-1996\\

@@

SCBI & JUNI & 30 & 30 & 28 & 28 & 20.4-76.19 & 5.6-59.5 & 1932-2009\\

@@

SCBI & LITU & 106 & 106 & 104 & 104 & 10-91.42 & 0.4-81.1 & 1932-2009\\

@@

SCBI & PIST & 36 & 36 & 36 & 36 & 13.92-50.96 & 0.5-44.3 & 1932-2009\\

@@

SCBI & QUAL & 66 & 66 & 66 & 66 & 11.4-76.73 & 0.3-70.4 & 1932-2009\\

@@

SCBI & QUMO & 67 & 67 & 67 & 67 & 10.22-84.59 & 1.4-69.5 & 1932-2009\\

@@

SCBI & QURU & 70 & 70 & 70 & 70 & 11.07-87.65 & 2.5-79.2 & 1932-2009\\

@@

SCBI & QUVE & 81 & 81 & 81 & 81 & 16.02-82.33 & 2-78.4 & 1932-2009\\

@@

SC & PIMA & 442 & 442 & 101 & 101 & 7-14.9 & 0.5-12.5 & 1927-2012\\

@@

ZOF & ABAL & 46 & 46 & 41 & 41 & 50-121 & 21.1-107.4 & 1903-2008\\

@@

ZOF & FASY & 1367 & 1367 & 1356 & 1356 & NA-NA & 0.1-115.3 & 1903-2008\\

@@

ZOF & PCAB & 640 & 640 & 596 & 596 & NA-NA & 0.5-126.4 & 1903-2008\\

@@

\hypertarget{table-s4.-allometric-equations-for-bark-thickness.}{%
\subsection{Table S4. Allometric equations for bark
thickness.}\label{table-s4.-allometric-equations-for-bark-thickness.}}

@@

\em{Abies alba} & $bark.mm = ((0.05 + 0.06 * (dbh.cm/2.54))/2)*2.54$ & NA &  & North America & Miles, Patrick D.; Smith, W. Brad. 2009.\\

@@

\em{Acer pseudoplatanus} & $bark.mm = 0.619 * log(dbh.cm + 1)$ & 10 & 8.2-39.6 & SCBI & Anderson-Teixeira et al. (2015)\\

@@

\em{Carya glabra} & $bark.mm = 1.035 * log(dbh.cm + 1)$ & 8 & 19.1-78 & SCBI & Anderson-Teixeira et al. (2015)\\

@@
